---
title: "Building functions"
author: "Yuka Takemon"
date: "14/06/2020"
slug: rstudio-training-ggplot2-tutorial
categories:
  - R
  - functions
  - tutorial
  - RLadies
tags: [R, functions, tutorial]
subtitle: ''
summary: ''
authors: [YukaTakemon]
lastmod: '2020-06-14'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

<br>

**Please note: This page is still under construction!**

<br>


This tutorial on "Building functions" was built for a workshop [R-Ladies Vancouver](https://twitter.com/RLadiesVan?ref_src=twsrc%5Egoogle%7Ctwcamp%5Eserp%7Ctwgr%5Eauthor) and adapted works from [The Carpentries](http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2/index.html) licensed under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). 

## Objectives

In this short tutorial learners will understand:

* What functions are and how are they useful
* Why custom functions are useful and when we might need them
* How to build your own basic and complex functions

***

## What are functions?

Functions packages a sequence of operations into one, preserving it for ongoing use without the need to repeat them. In R functions are invoked with a special term followed immediately by `()`. For example you've probably already come across functions such as `library()` to load packages, `print()` to print characters, and `paste()` to string together multiple vectors to create one long vector. 

You have probably noticed that functions provide:

* A single name to invoke operations
* A relief from the need to remember the individual operations
* A defined set of inputs and expected outputs

Functions are the basic building blocks of most programming languages, and user-defined functions constitute what we call “programming”. **If you have written a function, you are a computer programmer!**

***

## Why build custom functions?

Although base R package and packages on CRAN provide many useful functions, there are instances where you'll need to repeat a series of operations that you wish could be summed up into one function. 

***

## What are functions?

Before we talk about details let's build our first function together and try using it. 

```{r}
multiply <- function(a, b) {
  the_product <- a * b
  return(the_product)
}

multiply(3,4)
```

Here are the bare bones essential components of a function

* `variable <- function()`
* arguments within `()`
* series of operations that leads to a result
* `return()` result

```{r, eval = FALSE}
variable <- function(arguments) {
  result <- operations
  return(result)
}
```


You may have noticed a few things:

* no verbose

### <span style="color: darkorange;">Challenge 1</span>:
Build a function that 


```{r}

```

