---
title: "Building functions"
author: "Yuka Takemon"
slug: rladies-building-functions
categories:
  - R
  - functions
  - tutorial
  - RLadies
  - bonus_content
  - emoji
tags: [RLadies, R, functions, tutorial, bonus_content, emoji]
subtitle: ''
summary: ''
authors: [YukaTakemon]
lastmod: '2020-06-21'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
library(emo)
```

**Please note: This page is still under construction!**

* Pending code of conduct

## Objectives

This short tutorial will attempt to show:

* What functions are and how are they useful
* Why custom functions are necessary and when we might need them
* How to build your own basic functions and how to source them

This tutorial on "Building functions" was built for a [R-Ladies Vancouver](https://twitter.com/RLadiesVan?ref_src=twsrc%5Egoogle%7Ctwcamp%5Eserp%7Ctwgr%5Eauthor) workshop, which adapted works from [The Carpentries](http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2/index.html) licensed under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). 

***

## What are functions?

Functions package a sequence of operations into one, preserving it for ongoing use without the need to repeat them. In R functions are invoked with a special term followed immediately by `()`. For example you've probably already come across functions such as `library()` to load packages, `print()` to print characters, and `paste()` to string together multiple vectors to create one long vector. 

You have probably noticed that functions provide:

* A single memorable name to invoke operations
* A relief from the need to remember the individual operations
* A defined set of inputs and expected outputs

Functions are the basic building blocks of most programming languages, and user-defined functions constitute what we call “programming”. **If you have written a function, you are a computer programmer!**

***

## Why build custom functions?

Although base R package and packages on CRAN provide many useful functions, there are instances where you'll need to repeat a series of operations that you wish could be summed up into one function. 

***

## Defining functions

**Here are the essential building blocks of a function**

* `meaningful_name <- function()`
* arguments are defined within `()`
* series of operations are inclused within `{}`
* `return()` result

```{r, eval = FALSE}
meaningful_name <- function(argument1, arugment2, ...) {
  result <- operation_using(argument1, argument2)
  return(result)
}
```

Let's build our first function together and try it out. 

```{r}
multiply <- function(a, b) {
  the_product <- a * b
  return(the_product)
}

multiply(3, 4)
```

### <span style="color: lightgreen;">What did you observe?</span>:

What happened when you executed `multiply <- function(a, b){...}`?

* No verbose
* Addition of function to _global environment_
* No intermediate variables produced in _global environment_

### <span style="color: darkorange;">Challenge 1:</span> (10 mins)
Using the building blocks I showed you above, fill in the blanks below to build function that converts Fahrenheit to Kelvin. Here is the formula: `(°F − 32) × 5/9 + 273.15 = K`.

```{r, eval = F}
F_to_K <- _____(temp) {
  kelvin <- ________
  _____(kelvin)
}
```

Let's test it out! 32°F is the freezing point, which should correspond to 273.15K.
```{r, echo = F}
F_to_K <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

```{r}
F_to_K(32)
```

***

### <span style="color: darkorange;">Challenge 2:</span> (10 mins)
Now let's build a function called `K_to_C()` that converts Kelvin to Celsius. The formula for the conversion is: `K - 273.15 = Celsius`.

Take the next 10 mins to figure out which on of these will give us the right function? if not why?  Feel free to try them about.

A.
```{r, eval = F}
K_to_C() <- function(temp){
 celsius <- temp - 273.15
 return(celsius)
  }
```

B.
```{r, eval = F}
K_to_C <- function(temp){
 celsius <- temp - 273.15
  }
```

C.
```{r, eval = T}
K_to_C <- function(kelvin){
 celsius <- kelvin - 273.15
 return(celsius)
  }
```

D.
```{r, eval = F}
K_to_C <- function(kelvin){
 celsius <- temp - 273.15
 return(celsius)
  }
```

***

## Combining functions

The real power of programming comes when you can mix and match various functions together. Once functions are defined, we can use it within other functions or string multiple functions together to create a single more powerful operation

### <span style="color: darkorange;">Challenge 3:</span> (15 mins)

Build a new function `F_to_C()` that uses the functions we previously built, `F_to_K()` and `K_to_C()`, to convert Fahrenheit directly to Celsius. 

Let me start you off:

```{r}
F_to_C <- function(){
  
}

```

***

## Sourcing functions

There are different ways to call on custom defined function. As we have been doing, you can defined your custom function at the beginning of each script as you use them. However, if you are using the same custom functions across different projects scripts it becomes too repetitive and forces you to either remember each step and dig through previous scripts to copy and paste it into your new script.

In programming, we want to reduce repetition if possible, and this is where sourcing functions come in. 

First, let's open a new R script and paste our custom functions from below. Next save the the code block in to a script called. `my_func.R`. 

```{r}
F_to_K <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}

K_to_C <- function(kelvin){
 celsius <- kelvin - 273.15
 return(celsius)
}

F_to_C <- function(temp){
  K_result <- F_to_K(temp)
  celsius <- K_to_C(K_result)
  return(celsius)
}
```

Now let's delete all the variables and functions we've created so far and "reset" our environment using `rm(list = ls())`

To source our set of three functions, we use the `source()` function locate our `my_func.R` script.

```{r, eval = F}
source("~/Desktop/my_func.R")
```

Do you see three familiar functions in your _global environment_?

***

## Concluding thoughts

* **Congratulations!** you are now a programmer, huzzah! `r ji("partying_face")` `r ji("tada")` 
* Custom functions can be useful for repetitive operations 
* Functions can be defined per script or called upon using `source()`

***

## Bonus content: Emojis

While making this lesson, I learn to add Emojis into a R markdown document. I thought I'd share that here `r ji("wink")`:

You'll need to install the [Emo package](https://github.com/hadley/emo) from GitHub developed by [Hadley Wickham](https://github.com/hadley/)

```{r, eval= F}
install.packages("devtools")
devtools::install_github("hadley/emo")
library(emo)
```

There are two ways to print emojis by using:
* `ji_keywords`
* `ji`

For example, `ji_keywords` evoke none unique emojis that call on similar category of emo(ji)s. For example `cats` have various unique cat expressions:


```{r}
keywords <- ji_keyword
keywords$cat
```

```{r, echo = F}
cat_cats <- NULL
for(i in keywords$cat){
  print(ji(i))
}
```

To specify which specific cat you want you'll have to specify that cat you want using `ji()` like so: 

```{r}
ji("heart_eyes_cat")
```

You can find all the keywords on the [Emo package page](https://github.com/hadley/emo) or see below for a sample:

```{r, echo = F}
keywords <- keywords[lengths(keywords) > 1]
emojis <- purrr::map_chr(keywords[20:30], function(x) paste0(emo::ji_name[x], collapse = ""))
cat(paste(names(emojis), ": ", emojis, collapse = "\n"))
```

***

**Have suggestions or comments? [Contact me](https://yukatakemon.netlify.app/#contact)**

