---
title: "IC50 calculations using tidypredict"
author: Yuka Takemon
date: '2020-03-22'
slug: tidymodel-ic50-tutorial
categories:
  - R
  - tidymodel
  - tutorial
tags: [R, tidymodel, tutorial]
subtitle: ''
summary: ''
authors: [YukaTakemon]
lastmod: '2020-03-22'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

## Why I decided to write this
I needed to calculate the IC50 of a drug and at first glance it looked like most folks use pre-built tools on [PRISM](https://www.graphpad.com/scientific-software/prism/). I haven't yet taken the time to explore [tidymodels](https://github.com/tidymodels), so I decided to take this opportunity to learn some basic features. Also I have always been a big fan of open access & equal opportunity, especially when it comes to data analyses. Hopefully, a few people will stumble on to this and find it helpful for their work.  

***

## What is IC50?

IC50, half maximal inhibitory concentration, is a measure of how much inhibitory substance is needed to inhibit 50% of a given population. Frequently IC50 is calculated to determine the concentration/dose of drug necessary to have a survival of 50% _in vitro_.

Here we use [tidypredict](https://github.com/tidymodels/tidypredict), a package that is apart of the [tidymodels](https://github.com/tidymodels) universe, to create a simple function called `ic50` to predict the IC50 given a dosage response curve. 

***

## Modeling 

We will take `mtcars` to create a fake dosage response data set to fit a linear model. We will reassign disp as inhibition data and wt as dosage. 
```{r}
library(pacman)
p_load(tidypredict, tidyverse)
  
df <- mtcars %>% 
  select(disp, wt) %>%
  rename(inhibition = disp,
         dose = wt)

glimpse(df)
```

Since we know the maximum and minimum dosage and inhibition, we can calculate the 50 percentile inhibition. This will be used to predict the dosage necessary to achieve an IC50. 
```{r}
ic50 <- function(df){
  
  # Generate linear model of based on simulated dose response.
  model <- lm(dose ~ inhibition, data = df)
  
  # Get 50th-percentile inhibition
  test <- tibble(inhibition = max(df$inhibition) - (max(df$inhibition) - min(df$inhibition))/2)
 
  return(tidypredict_to_column(test, model, vars = "dose")) # print new column "dose" and append prediction
}

ic50(df)
```

We calculate the 50% inhibition, `272`, which we predict to be achieved using a dosage of 3.5. 

Let's plot the data to see what our hypothetical data would look like and how our prediction stacks up:
```{r}
pred <- ic50(df)

ggplot(df, aes(dose, inhibition))+
  geom_point()+
  geom_point(data = pred, aes(x = dose, y = inhibition), 
             colour = "red", size = 5)+
  geom_smooth(method = "lm", se = F)
```

Here we show how our IC50 prediction stacks up in our dosage response data we simulated. At this point I would be comfortable using a dosage of 3.5 moving forward.

***

